---
title: "Collected/Imputed vs FLU 2016"
date: "11/9/2020"
output: 
  html_notebook:
    theme: readable
    toc: TRUE
    toc_float: true 

---
# Data Notes
```{r}
datafile <- "J:/Staff/Christy/usim-baseyear/flu/impute_flu_comparison_filled_2020-11-04.csv"
```

Analysis comparing prev FLU (2016) to what has been collected or imputed.

This analysis uses `r datafile`.

Notes about the data:

* Imputation is based on the following formulas
  - $DU = (max height/15)^2$
  - $FAR = max height/20$
* Columns ending in `_imp` contain both imputed, collected, or previous density values
* Columns ending in `_src` indicate whether value was imputed, collected, or from the previous FLU

# Tidy Data
### Transform Data
```{r prep data}
library(data.table)
library(tidyverse)
library(patchwork)
library(DT)
library(treemapify)

df <- fread(datafile)

# create long format 
id.cols <- c("Key", "Jurisdicti_new", "Definition_new")  
prev.cols <- str_subset(colnames(df), "_prev")
msr.cols <- str_subset(colnames(df), "_imp")
src.cols <- str_subset(colnames(df), "_src")

dfm.src <- melt(df, id.vars = c(id.cols, msr.cols), measure.vars = src.cols, variable.name = "source_type", value.name = "source_value", variable.factor = F)

dfm <- melt(dfm.src, id.vars = c(id.cols, "source_type", "source_value"), measure.vars = msr.cols, variable.name = "imputed_type", value.name = "imputed_value", variable.factor = F)

# remove extraneous records created because of melts
dfm[, `:=` (keyimp = str_extract(imputed_type, "^\\w+_\\w+(?=_\\w+)"),
            keysrc = str_extract(source_type, "^\\w+_\\w+(?=_\\w+)"))]
dfm <- dfm[keyimp == keysrc, ]

# subset for 2016 columns, remove duplicate rows (b/c orig. format is shapefile)
sub.cols <- c("Key", prev.cols)
dfs <- df[, ..sub.cols]
dfu <- unique(dfs) # remove duplicate rows

# join melted df to prev df and add type columns
dfj <- dfm[dfu, on = "Key"]
dfj[, `:=`(dens_type = str_extract(imputed_type, "^Max\\w{2,3}(?=_)"),
             use_type = str_extract(imputed_type, "(?<=_)\\w+(?=_)"))]

# add dens type to criteria
dfj[!is.na(Max_FAR_prev) & !is.na(imputed_value) & dens_type == "MaxFAR", max_far_diff := imputed_value-Max_FAR_prev]
dfj[!is.na(Max_DU_Ac_prev) & !is.na(imputed_value) & dens_type == "MaxDU", max_du_diff := imputed_value- Max_DU_Ac_prev]
dfj[source_value == "", source_value := NA]
dfj <- dfj[!is.na(source_value)]

# Height data
ht.cols <- str_subset(colnames(df), "MaxHt.*")
ht.df.cols <- c(id.cols, ht.cols, "Max_Height_prev") 
dfh <- df[, ..ht.df.cols]

dfm.ht <- melt(dfh, id.vars = c(id.cols, "Max_Height_prev"), measure.vars = ht.cols, variable.name = "height_type", value.name = "height", variable.factor = F)
dfm.ht[, `:=` (use_type = str_extract(height_type, "(?<=_)\\w+$"),
               height_diff = height - Max_Height_prev)]

```

### Functions
```{r ggplot functions}
scatterplot <- function(df, xvar, yvar, colorvar, srcvar) {
    ggplot(df) +
    geom_point(aes_string(x = xvar, y = yvar, color = colorvar, shape = srcvar),
             alpha = .7) +
    scale_shape_manual(values = c(1,2,3,4,5)) +
    geom_abline()
}

scatterplot_facet <- function(df, xvar, yvar, colorvar, srcvar, rowfacet, colfacet) {
  scatterplot(df, xvar, yvar, colorvar, srcvar) +
    facet_grid(rows = vars(.data[[rowfacet]]), cols = vars(.data[[colfacet]]))
}

dotplot <- function(df, xvar) {
  ggplot(df) +
    geom_dotplot(aes_string(xvar), dotsize = .6) +
    scale_y_continuous(NULL, breaks = NULL)
}

dotplot_facet <- function(df, xvar, colfacet, rowfacet) {
  dotplot(df, xvar) +
    facet_grid(rows = vars(.data[[rowfacet]]), cols = vars(.data[[colfacet]]))
}

treemap <- function(df, areavar, fillvar, labvar, subgrpvar, colfacet) {
  ggplot(df, aes_string(area = areavar, fill = fillvar, label = labvar, subgroup = subgrpvar), guide = F) +
  geom_treemap() +
  geom_treemap_text(min.size = 1, reflow = T) +
  geom_treemap_subgroup_text(color = "white", alpha = .4) +
  facet_wrap( ~ .data[[colfacet]]) +
  guides(fill = F)
}

log_scale <- list(scale_y_log10(), scale_x_log10())

```
# Visuals
## Max Densities
### Current vs Previous
```{r, fig.height=5, fig.width=10}
du <- scatterplot(dfj[dens_type == "MaxDU", ], "Max_DU_Ac_prev", "imputed_value", "use_type", "source_value") +
  labs(title = "Default scale")
du_log <- du + log_scale + labs(title = "Log scale")
du + du_log & labs(y = "Current density")
```



```{r, fig.width = 10}
far <- scatterplot(dfj[dens_type == "MaxFAR", ], "Max_FAR_prev", "imputed_value", "use_type", "source_value")
far + labs(y = "Current Density")
```
### Density Difference Ranges
```{r, fig.height=8, fig.width=10}
df.du <- dfj[dens_type == "MaxDU"]
df.far <- dfj[dens_type == "MaxFAR"]

far_diff_dotplot <- dotplot_facet(df.far, "max_far_diff", "use_type", "source_value")

far_diff_dotplot + labs(title = "Max FAR difference by Use and Source Type")
```
```{r, fig.height=8, fig.width=10}
du_diff_dotplot <- dotplot_facet(df.du, "max_du_diff", "use_type", "source_value")

du_diff_dotplot + labs(title = "Max DU difference by Use and Source Type")

```
### Density Difference Outliers 

```{r, fig.width=10}
du.outliers <- boxplot.stats(df.du$max_du_diff)
du.out <- data.frame(outliers = du.outliers$out)

far.outliers <- boxplot.stats(df.far$max_far_diff)
far.out <- data.frame(outliers = far.outliers$out)

far_diff_out_dotplot <- dotplot(far.out, "outliers") + labs(x = "Max FAR difference", title = "MaxFAR outliers")
du_diff_out_dotplot <- dotplot(du.out, "outliers") + labs(x = "Max DU difference", title = "MaxDU outliers")

far_diff_out_dotplot + du_diff_out_dotplot
```

```{r, fig.width = 10, fig.width=10}
df.du.out <- df.du[max_du_diff %in% du.outliers$out]

ggplot() +
  geom_point(data = df.du.out, aes(Max_DU_Ac_prev, imputed_value, color = use_type, shape = source_value)) +
  geom_abline(color = "grey") +
  facet_wrap(Jurisdicti_new ~ .) +
  labs(title = "MaxDU outliers")

```
```{r, fig.width=10}
df.out.cols <- c("Jurisdicti_new", "Definition_new", "use_type")
df.du.out.grp <- df.du.out[, .(count = .N), by = df.out.cols]

# treemap(df.du.out.grp, "count", "Jurisdicti_new", "Definition_new", "Jurisdicti_new", "use_type") +
#   labs(title = "Zone Descriptions of MaxDU Outliers by Use Type and Jurisdiction")

treemap(df.du.out.grp, "count", "use_type", "Definition_new", "use_type", "Jurisdicti_new") +
  labs(title = "Zone Descriptions of MaxDU Outliers by Use Type and Jurisdiction")
```


```{r, fig.width=10}
df.far.out <- df.far[max_far_diff %in% far.outliers$out]

ggplot() +
  geom_point(data = df.far.out, aes(Max_FAR_prev, imputed_value, color = use_type, shape = source_value)) +
  geom_abline(color = "grey") +
  facet_wrap(Jurisdicti_new ~ .) +
  labs(title = "MaxFAR outliers") +
  theme(axis.text.x = element_text(size = 7, angle = 45, hjust = 1, vjust = 1),
        axis.text.y = element_text(size = 8))
```
```{r, fig.width=10}
df.far.out.grp <- df.far.out[, .(count = .N), by = df.out.cols]

# treemap(df.far.out.grp, "count", "Jurisdicti_new", "Definition_new", "Jurisdicti_new", "use_type") +
#   labs(title = "Zone Descriptions of MaxFAR Outliers by Use Type and Jurisdiction")

treemap(df.far.out.grp, "count", "use_type", "Definition_new", "use_type", "Jurisdicti_new") +
  labs(title = "Zone Descriptions of MaxFAR Outliers by Use Type and Jurisdiction")
```
## Height
### Current vs Previous
```{r, fig.width=10}
ht <- scatterplot(dfm.ht, 'Max_Height_prev', "height","use_type", NULL) + 
  labs(y = "Current Height", title = "Default Scale")
ht.log <- scatterplot(dfm.ht, 'Max_Height_prev', "height","use_type", NULL) + 
  log_scale +
  labs(title = "Log Scale")
ht + ht.log
```

### Height Difference Ranges

```{r}
dotplot(dfm.ht, "height_diff") + labs(title = "Range of Height Differences")
```
### Height Difference Outliers
```{r}
ht.outliers <- boxplot.stats(dfm.ht$height_diff)

ht.out <- dfm.ht[height_diff %in% ht.outliers$out]

dotplot(ht.out, "height_diff") + labs(title = "Height Difference Outliers")

```

```{r}


ggplot() +
  geom_point(data = ht.out, aes(Max_Height_prev, height, color = use_type)) +
  geom_abline(color = "grey") +
  facet_wrap(Jurisdicti_new ~ .) +
  labs(title = "Max Height outliers") +
  theme(axis.text.x = element_text(size = 7, angle = 45, hjust = 1, vjust = 1),
        axis.text.y = element_text(size = 8))

```

```{r, fig.width=10}
df.ht.out.grp <- ht.out[, .(count = .N), by = df.out.cols]

# treemap(df.ht.out.grp, "count", "Jurisdicti_new", "Definition_new", "Jurisdicti_new", "use_type") +
#   labs(title = "Zone Descriptions of Max Height Outliers by Use Type and Jurisdiction")

treemap(df.ht.out.grp, "count", "use_type", "Definition_new", "use_type", "Jurisdicti_new") +
  labs(title = "Zone Descriptions of Max Height Outliers by Use Type and Jurisdiction")
```


