---
title: "Matching Buildings to OFM"
author: "Hana Sevcikova"
date: "3/11/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Setup

#### Code and Data

This R markdown file is located in the [2018 imputation GitHub repo]( https://github.com/psrc/urbansim-baseyear-prep/blob/master/imputation/imp2018/ofm_matching.Rmd).

Input files should be placed in the [imputation/data2018]( https://github.com/psrc/urbansim-baseyear-prep/blob/master/imputation/data2018) directory. These are:

* imputed_buildings.csv (output of [buildings_imputation.Rmd]( https://github.com/psrc/urbansim-baseyear-prep/blob/master/imputation/imp2018/buildings_imputation.Rmd))
* parcels.csv
* OFMPopHHblocks.csv (containing the OFM data)
* census_blocks.csv

#### Loading Inputs

In this processing, input files are taken from the directory
```{r, echo=FALSE}
library(data.table)
data.year <- 2018 # data files will be taken from "../data{data.year}"
data.dir <- file.path("..", paste0("data", data.year))
normalizePath(data.dir)
```


```{r, echo=FALSE}
# read buildings and parcels table
bld.file.name <- 'imputed_buildings.csv'
bld.imp <- fread(file.path(data.dir, bld.file.name))
pcl <- fread(file.path(data.dir, "parcels.csv"))
setkey(pcl, parcel_id)

# reads OFM DU totals from the data directory (creates object ofm)
source(file.path(data.dir, "load_ofm.R"))
```

```{r, echo=FALSE}
# join buildings and parcels
bld <- copy(bld.imp)
setkey(bld, building_id, parcel_id)
bld[, county_id := NULL] # will be taken from parcels dataset
bld <- merge(bld, pcl[, .(parcel_id, census_block_id, census_tract_id)], by = "parcel_id")
# assign county_id from census_blocks (since the parcel.county_id might have issues)
cb <- fread(file.path(data.dir, "census_blocks.csv"))
bld <- merge(bld, cb[, .(census_block_id, county_id)], by = "census_block_id")
```

The buildings table `r bld.file.name` has been merged with parcels and has `r nrow(bld)` rows. The OFM dataset has `r nrow(ofm)` records. 

#### Settings

Geographies by which we want to match the datasets:

```{r, echo=TRUE}
match.by <- c("county_id", "census_tract_id", "census_block_group_id"
              #, "census_block_id"
              )
```

Set building types to be considered as residential and if the block group dataset should be written into a file.
```{r, echo=TRUE}
allrestypes <- c(12, 4, 19, 11, 10)
save.block.groups <- FALSE
```


```{r, echo=FALSE}
if(save.block.groups && ! "census_block_group_id" %in% match.by)
  stop("Either census_block_group_id should be included in match.by or save.block.groups should be set to FALSE")
```


```{r, echo=FALSE}
mftypes <- c(12, 4) # MF building type 

allrestypes <- c(12, 4, 19, 11, 10) # All residential building types

gqbtypes <- 6 # group-quarters building types
            
# Residential land use types
reslutypes <- c(13, # mobile home
                14, # MF
                15, # condo
                21, # recreation
                24, # SF
                26, # vacant developable
                30 # mix use
                )

# Group-quaters land use types
gqlutypes <- c(8,  # GQ
               11  # military
               )
```

### Processing

```{r, echo=FALSE}
# aggregate developable blocks
dev.blocks <- pcl[, .(Nrespcl = sum(land_use_type_id %in% reslutypes), 
                      Ndevpcl = sum(plan_type_id != 1000),
                      Ngqpcl = sum(land_use_type_id %in% gqlutypes),
                      Npcl = .N), 
                  by = match.by]

# reduce ofm to only needed attributes
ofm <- ofm[, c(match.by, "HU", "GQ"), with = FALSE]

# Function for merging blds and OFM 
merge.du.ofm <- function() {
	du <- bld[, list(
	        DU=sum(residential_units), 
	        number_of_buildings=.N, 
	        Nmf=sum(building_type_id %in% mftypes),
					Nres=sum(building_type_id %in% allrestypes),
					Ngq=sum(building_type_id %in% gqbtypes),
					DUimp=sum((residential_units - residential_units_orig) * 
					            (residential_units > residential_units_orig) * 
					            (building_type_id %in% allrestypes)),
					DUred=sum((residential_units_orig - residential_units) * 
					            (residential_units < residential_units_orig) * 
					            (building_type_id %in% allrestypes))
					), by=match.by]
	aggofm <- ofm[, .(HU = sum(HU), GQ = sum(GQ)), by = match.by]
	# merge with OFM totals (created in load_ofm.R)
	duofm <- merge(du, aggofm, by=match.by, all = TRUE)
	duofm <- merge(duofm, dev.blocks, by=match.by, all = TRUE)
	# remove blocks with zero OFM and no buildings
	duofm <- duofm[!(is.na(DU) & HU == 0)] 
	# groups with no buildings in our dataset - set attributes to 0
	duofm[is.na(DU), no_buildings := TRUE]
	duofm[is.na(no_buildings), no_buildings := FALSE]
	duofm[no_buildings == TRUE, 
	      c("DU", "number_of_buildings", "Nmf", "Nres", "DUimp", "DUred")] <- 0
	# groups with no OFM records
	duofm[is.na(HU), no_ofm := TRUE]
	duofm[is.na(no_ofm), no_ofm := FALSE]
	duofm[no_ofm == TRUE, c("HU", "GQ")] <- 0
	# compute difference between us and OFM
	duofm[, dif := DU - HU]
	# set various attributes to 0
	for(attr in c("Nrespcl", "Ndevpcl", "Ngq", "Ngqpcl", "Npcl"))
		duofm[is.na(duofm[[attr]]), attr] <- 0
  # GQ difference as boolean
	duofm[, gqdif := (Ngq > 0) - (GQ > 0)]
	duofm
}

```

```{r, echo=FALSE}
duofm <- merge.du.ofm()
```

When merging buildings and OFM, the following was found:

* `r sum(duofm$no_ofm)` groups contain buildings (with `r sum(duofm[no_ofm == TRUE, DU])` residential units in total) but no OFM records. 
    * These are: `r paste(match.by[length(match.by)], paste(sort(unique(duofm[no_ofm == TRUE,][[match.by[length(match.by)]]])), collapse = ", "), sep = ":")`
* `r sum(duofm$no_buildings)` groups that have OFM records (with `r as.integer(sum(duofm[no_buildings == TRUE, HU]))` DUs in total) but no buildings.
* `r nrow(duofm[HU == 0 & DU > 0])` groups for which there are zero OFM units and we have `r duofm[HU == 0 & DU > 0, sum(DU)]` residential units.
* `r nrow(duofm[DU == 0 & HU > 0])` groups for which we have zero residential units and there are `r as.integer(sum(duofm[DU == 0 & HU > 0, HU]))` OFM units.


### Results

In the following tables, the DU column represents residential units including imputation. Imputed units are shown in the column DUimp. Thus, the original value for residential units can be derived as DU - DUimp. Column DUofm represent the OFM counts.

#### Total

```{r, echo=FALSE}
aggregate.duofm <- function(dt, by = NULL) {
  dt[, .(DU=sum(DU), DUimp=sum(DUimp), DUofm=round(sum(HU)), dif=round(sum(dif))), by=by]
}
```

```{r, echo=FALSE}
aggregate.duofm(duofm)
```

#### By county

```{r, echo=FALSE}
aggregate.duofm(duofm, by = "county_id")
```

#### By block group

`r if("census_block_group_id" %in% match.by) "Here are block groups with the 15 largest positive differences:"`

```{r, echo=FALSE}
if("census_block_group_id" %in% match.by) {
  aggdt <- aggregate.duofm(duofm, by = c("county_id", "census_block_group_id"))
  head(aggdt[order(dif, decreasing = TRUE)], 15)
}
```

`r if("census_block_group_id" %in% match.by) "Here are block groups with the 15 largest negative differences:"`

```{r, echo=FALSE}
if("census_block_group_id" %in% match.by) 
  head(aggdt[order(dif)], 15)
```

`r if("census_block_group_id" %in% match.by) "Here is a scatter plot of DU ~ DUofm for all block groups:"`

```{r, echo=FALSE}
if("census_block_group_id" %in% match.by) {
  plot(DU ~ DUofm, data = aggdt)
  abline(0,1)
  sdat <- head(aggdt[order(abs(dif), decreasing = TRUE)], 10)
  text(DU ~ DUofm, data = sdat, labels = sdat$census_block_group_id, cex = 0.6, pos = 2)
}
```

`r if(!"census_block_group_id" %in% match.by) "No aggregation by block groups available. Include census_block_group_id into match.by."` 

```{r, echo=FALSE}
file.bg <- "ofmdif_bg.csv"
if(save.block.groups) {
  fwrite(aggdt, file = file.bg)
}
```

If `save.block.groups` is `TRUE`, the resulting file is written into `r  file.bg`. Currently, `save.block.groups` is `r save.block.groups`.
